---
title: "Student Wellness"
author: "Aman Panwar"
output: html_document
---

**1. Problem Context**

Physical inactivity has been responsible for six percent of the global mortality and it has been ranked as the fourth leading risk factor of death (World Health Organization, 2010). For the age group of 5???17 years, physical activity provides essential health benefits for children and youth (Janssen, 2007; Janssen & Leblanc, 2010; Physical Activity Guidelines Advisory Committee, 2008). In addition, the behaviors related to health during the early teenage years between 11 and 16 years are crucial because the patterns of the health-related habits in their adulthoods emerge at this time (Gregory & Lowe, 2000).
In this context, the importance of students' health and the necessity to figure out of the factors influencing the students' physical fitness have been emphasizing. Safe neighborhoods that are free of crime and violence are an integral component of healthy neighborhoods. In addition to direct physical and mental impacts of crime, fear of crime and violence inhibit the use of community assets  for physical exercise at nearby parks and playgrounds or walking or bicycling to commute to local destinations for basic needs (Fowler et al., 2009; Takagi, Ken'ichi, & Kawachi, 2012).  

In this project, we decided to use multivariate statistical techniques to obtain deeper insights into the relationship between the rate of crime of an area and physical fitness of students living there.


**Datasets:**

Two datasets:physical fitness and crime datasets were used in this project. 

Physical fitness dataset had physical fitness score for students who attended grades five, seven, and nine in California public schools from 1998 to 2018.

Crime dataset had the rate of crime which represents the number of violent crimes per 1,000 population. Four types of crimes (murder, rape, robbery, and assault) and the total number of crimes are reported from years 2000 to 2013 at the geographical levels of city/towns, counties, regions, and state.



Libraries to perform multivariate analysis

```{r, warning=FALSE,message=FALSE}
library(dplyr)
library(reshape2)
library(imp4p)
library(MVA)
library('ResourceSelection')
library('webshot')
```

**2. Data Cleaning**

Physical fitness dataset
```{r}
#Physical fitness data
p_fit = read.csv("pfitness.csv")
#Changes name of the county column to make column name consistent in both datasets
names(p_fit)[5] <- "county_name"

#Subsets data to include only necessary columns
p_fit_1  <- p_fit[, c("Year", "county_name", "Grade.Lev", "Percent", "Strata")]
head(p_fit_1)
```

```{r}
#The crime dataset had data for only 14 years so data for all years except 2000-2013 was removed from the physical fitness dataset.

#Removes unnecessary rows from Year columns to get data only for 14 years (2000-2013)
p_fit_2 <- p_fit_1[!(p_fit_1$Year=="1998-1999"|p_fit_1$Year=="1999-2000"
                     |p_fit_1$Year=="2014-2015"|p_fit_1$Year=="2015-2016"|p_fit_1$Year=="2016-2017"|p_fit_1$Year=="2017-2018"|    
                       p_fit_1$county_name == "California"|p_fit_1$county_name == "Ca Education Authority"),]

#Selects data for all students
p_fit_3 <- subset(p_fit_2, Strata=="None")

#Checks for missing values in each column
colSums(is.na(p_fit_3))
```

```{r}
#County column will be used to merge both datasets
#Groups data by county
p_fit_group <- group_by(p_fit_3,county_name)

#Calculates average physical fitness scores for all students from all grades and years and grouped the data by county
summ_fit <- summarize(p_fit_group, avg_fitness_score =mean(Percent))
head(summ_fit)
```

Crime dataset
```{r}
#Crime dataset
v_crime <- read.csv("vcrime-2.csv")

# Selects necessary columns
v_crime_2 <- v_crime[, c("reportyear", "county_name", "strata_level_name", "geotype", "rate")]

# Subsets data based on county
v_crime_3 <- subset(v_crime_2, geotype=="CO")

head(v_crime_3)
```

```{r}
#converts crime rate from char to numeric
v_crime_3$rate <- as.numeric(v_crime_3$rate)

#Checks for missing values in each column
colSums(is.na(v_crime_3))
```

```{r}
#groups data by county
v_crime_group <- group_by(v_crime_3,county_name,strata_level_name )
head(v_crime_group)
```
```{r}
#summarizes average crime rate for each county for time period between 2000-2013
summ_crime <- summarize(v_crime_group, avg_crime_rate= mean(rate)) 
head(summ_crime )
```
```{r}
#summarizes average crime rate for each type of crime for each county 
v_crime_4 <- dcast(summ_crime, county_name~strata_level_name, value.var = "avg_crime_rate")
# Change the name of the columns
colnames(v_crime_4) <- c("county_name", "Aggravated_assault", "Forcible_rape", "Murder",
                         "Robbery", "Violent_crime_total")
head(v_crime_4)
```

**Data merging**
```{r}
merged_data <- merge(summ_fit, v_crime_4, by="county_name")
#removes categorical column to allow replacement of missing values with median
merged_data.1 <- merged_data[, -1]

#Handling of Missing Values

# Missing value imputation with median
for(q in 1:ncol(merged_data.1)){
  merged_data.1[is.na(merged_data.1[,q]),q] <- median(merged_data.1[,q], na.rm=TRUE)
}

colSums(is.na(merged_data.1))
#No missing values in merged dataset
head(merged_data.1)
```


**3. Data Visualization**


```{r}
#Bivariate Boxplot to find the outliers
county <- merged_data$county_name
merged_data.2 <- cbind(county,merged_data.1)
bvbox(merged_data.2[, c("avg_fitness_score","Violent_crime_total")])
text(merged_data.2[, c("avg_fitness_score", "Violent_crime_total")], labels = abbreviate(merged_data$county_name), cex = 0.7)
```


The bivariate plot indicates that there were 5 outlier counties that needed to be removed to get more reliable results, which increases the reliability of drawn conclusions from this analysis. 

```{r}
#Removes outliers
outlier <- match(c("Alpine", "San Joaquin", "Imperial", "El Dorado", "Marin"), merged_data.2$county)

#Dataset without any outliers
merged_data.o <- merged_data.2[-outlier, ]

#Scatterplot with labels and rugs(scales)
plot(merged_data.o$avg_fitness_score, merged_data.o$Violent_crime_total)
rug(merged_data.o$avg_fitness_score, side=1)
rug(merged_data.o$Violent_crime_total, side = 2)
text(merged_data.o$avg_fitness_score, merged_data.o$Violent_crime_total, cex = 0.6, labels = abbreviate(merged_data.o$county))
```

**4. Dimension Reduction Analysis**

There are 6 continuous variables in our dataset, and it can be difficult if someone needs to derive insight from this dataset with several variables. To help overcome this problem, PCA (Principal Component Analysis) was performed to explain the dataset comprehensively with minimum possible variables.PCA was performed on the fitness-crime(merged) dataset with outliers removed from it.

```{r}
#Removes categorical column for PCA
merged_data.3 <- merged_data.o[,-1]
# Principal component analysis
merged_data.3_pca <- princomp(merged_data.3, cor = T) 
summary(merged_data.3_pca)
```

```{r}
#Biplot for PCA analysis
biplot(merged_data.3_pca, col=c("black", "red"), cex = 0.65)
```

```{r}
merged_data.3_pca$loadings
```

Interpretation of PCA:

The first two principal components (PC) represents 77% of the total variance and it is therefore enough to use first two PC to represent the original dataset. Based on their respective loadings, PC 1 can be thought to represent all crimes except Forcible_rape and Robbery and PC 2 can be thought to represent Forcible_rape and Robbery. A simple interpretation of PC1 score could be that county with higher PC1 score will have lower crime rates especially for the crimes with higher absolute magnitude of PC loadings and higher physical fitness level. 


**4. Cluster Analysis**

We performed cluster analysis for discovering all possible clusters of most possible homogeneous observations in our fitness-crime dataset.
```{r}
# Kdepairs plot for detecting clusters
kdepairs(merged_data.o[,2:7])
```
Based on KDE plot, there are one or two clusters in all our data.

We considered two clustering methods:hierarchical and K-means clustering.

In hierarchical clustering, data is not partitioned into particular groups in a single step. The process requires a series of partitions that can run from a single cluster containing all individuals to n clusters, each containing a single individual. This clustering method uses the distance matrix (standardized Euclidean distance) and is based on the choices of distance between groups further divided into three categories: single, complete and average.

```{r}
# Hierarchical clustering ---- Single, complete,average
merged_data.o_hc <- hclust(dist(merged_data.3), "single")
merged_data.o_hc <- hclust(dist(merged_data.3), "complete")
merged_data.o_hc <- hclust(dist(merged_data.3), "average")
```

```{r}
# Constructs scree plot to determine number of hierarchical clusters
merged_data.3.s <- dist(scale(merged_data.o[,-1]))
hc1 <- hclust(merged_data.3.s, "average")
plot(rev(hc1$height))
```
The screeplot shows that there are 2 clusters in our data.

```{r}
ct <- cutree(hc1, 2)
#plots the clustered data
plot(merged_data.o[, c(2,7)], col = ct, main = "HC Clusters") 
text( merged_data.o[, c(2,7)],labels =abbreviate(merged_data.o$county), col = ct,pch=" ")
```
```{r}
#K-means clustering 
#removes first column
merged_data.o.1=merged_data.o[,-1]

#scree plot
maxc = 20
plot.wgss = function(merged_data.o.1, maxc) {
  wss = numeric(maxc)
  for (i in 1:maxc)
    wss[i] = kmeans(merged_data.o.1, centers=i, nstart = 20)$tot.withinss
  plot(1:maxc, wss, type="b", xlab="Number of Clusters",
       ylab="Within groups sum of squares", main="Scree Plot")
}
plot.wgss(merged_data.o.1, 20)
```

```{r}

km <- kmeans(scale(merged_data.o.1), centers = 3)

plot(merged_data.o[, c(2,7)], col =km$cluster, main = "kmeans Clusters") 

```

We eventually concluded that hierarchical based clustering method is the best method to segregate our observations into two clusters because it was evident in its plot that there was clear separation between the observations from two clusters. Although, in the absence of information on true clusters, it is difficult to be completely confident if the data has been well segregated.



**Supporting Information:**

# Percentage of Physically-fit Students (LGHC Indicator)
# https://healthdata.gov/dataset/percentage-physically-fit-students-lghc-indicator

# Physical Fitness Testing (PFT)
# https://www.cde.ca.gov/ta/tg/pf/

# Student Poverty FRPM Data
# https://www.cde.ca.gov/ds/sd/sd/filessp.asp

# Public Schools and Districts Data Files
# https://www.cde.ca.gov/ds/si/ds/pubschls.asp

# Violent Crime Rate
# https://data.ca.gov/dataset/violent-crime-rate


**References:**
●  Fowler, P. J. et al. (2009). Community violence: A meta-analysis on the effect of exposure and mental health outcomes of children and adolescents. Dev Psychopathol, 21(1), 227−259.
● Takagi, D., Ken'ichi, I., & Kawachi, I. (2012). Neighborhood social capital and crime victimization: Comparison of spatial regression analysis. Soc Sci Med. 75(10), 1895−902.





